---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/HUI950319/BulkS4/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/HUI950319/BulkS4/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![pkgdown](https://github.com/HUI950319/BulkS4/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/HUI950319/BulkS4/actions/workflows/pkgdown.yaml)
<!-- badges: end -->

# BulkS4: S4 Classes for Bulk RNA-seq Data Analysis

**BulkS4** is a comprehensive R package that provides S4 classes and methods for bulk RNA-seq data analysis. The package offers a unified framework for storing, analyzing, and visualizing bulk RNA-seq data with integrated support for:

- 🧬 **Differential Expression Analysis** using DESeq2, edgeR, and limma-voom
- 📊 **Gene Set Enrichment Analysis (GSEA)**
- 🎯 **Gene Set Variation Analysis (GSVA)**
- 📈 **Rich Visualization Tools** for publication-ready plots
- 🗂️ **Integrated Data Management** with S4 classes

Find out more at https://hui950319.github.io/BulkS4/.

## ✨ Key Features

- **BulkRNAseq S4 Class**: Unified data structure for storing raw counts, normalized data, sample metadata, and analysis results
- **Data Validation**: Built-in data consistency checks and validation mechanisms
- **Flexible Constructor**: Support for creating objects from matrices or preprocessed lists
- **Rich Visualization**: Complete set of plotting functions for expression data, heatmaps, volcano plots, and more
- **MSigDB Integration**: Automatic loading of MSigDB gene sets (Hallmark, KEGG, Reactome, GO)
- **Multiple DE Methods**: Integrated support for DESeq2, edgeR, and limma-voom with method comparison
- **Gene Set Analysis**: GSEA and GSVA analysis capabilities with multiple gene set collections

## 📦 Installation

```{r, eval = FALSE}
# Install from GitHub (recommended)
if (!require("devtools")) install.packages("devtools")
devtools::install_github("HUI950319/BulkS4")

# Load the package
library(BulkS4)
```

## 🚀 Quick Start

### Load Example Data

BulkS4 comes with example datasets that you can use to get started:

```{r load-data}
library(BulkS4)

# Load example datasets included in the package
data(counts)      # Example RNA-seq count matrix (36888 genes × 6 samples)
data(metadata)    # Example sample metadata
data(geneSet_msig) # MSigDB gene sets

# Examine the example data
head(counts)
head(metadata)
```

### Create BulkRNAseq Object

```{r create-object}
# Create BulkRNAseq object from example data
bulk_obj <- BulkRNAseq(counts, metadata)

# View object summary
bulk_obj                     
```


## 🔬 Differential Expression Analysis

### Run Differential Analysis

```{r differential-analysis, eval=FALSE}
# Run differential expression analysis using DESeq2--add normalized data and diff results
bulk_obj <- runDiffAnalysis(bulk_obj, method = "DESeq2")
# View the updated object
bulk_obj    # View the updated object 
# Get differential expression results
diff_results <- getDiffResults(bulk_obj)
head(diff_results[[1]])
```


## 📊 Visualization Examples

BulkS4 provides comprehensive visualization functions for RNA-seq data analysis:

### Gene Expression Bar/Box Plots

```{r gene-expression-plots, fig.width=6, fig.height=5}
# Single gene expression
bar(bulk_obj, genes = "CD36", group = "group")
```

```{r multiple-genes-plot, fig.width=10, fig.height=6}
# Multiple genes expression
bar(bulk_obj, genes = c("CD36", "GAPDH", "ACTB"), group = "group")
```

### Heatmaps

```{r heatmap, fig.width=8, fig.height=6}

# Heatmap of specific genes
selected_genes <-c("TSPAN6", "TNMD", "DPM1", "SCYL3", "C1orf112", "FGR", "CFH")
hmap(bulk_obj, genes = selected_genes, group = "group", 
     show_rownames = TRUE, scale = "row")

# Heatmap of top 20 DEGs
hmap(bulk_obj, genes = 20, group = "group", 
     show_rownames = TRUE, scale = "row")

```

### PCA and Hierarchical Clustering

```{r pca-clustering, fig.width=10, fig.height=5}
# PCA and hierarchical clustering combined plot
# Note: This requires additional packages (factoextra, tinyarray, ggsci, patchwork)
pca_hc(bulk_obj)
```

### Volcano Plot

```{r volcano-plot, fig.width=6, fig.height=5 }
# Create volcano plot for differential expression results
volcano(bulk_obj, data = "DESeq2_treat_vs_con")
```

### MA Plot

```{r ma-plot, fig.width=6, fig.height=5 }
# Create MA plot
MA(bulk_obj, data = "DESeq2_treat_vs_con")
```

## 🎯 Gene Set Analysis

### Gene Set Enrichment Analysis (GSEA)

```{r gsea-analysis, fig.width=8, fig.height=5}
# Run GSEA analysis-- use genesets from MSigDB
bulk_obj <- gsea(bulk_obj, geneSet_name = "H" )

# # Run GSEA analysis with common used genesets
# bulk_obj <- gsea(bulk_obj, geneSet_name = c("H", "C2_CP:KEGG_LEGACY", "C5_GO:BP", "C5_GO:CC", "C5_GO:MF") )

# View bulk_obj--gsea results add to gsea slot
bulk_obj
# head(bulk_obj@gsea$H)

# Visualize GSEA results
dotplot(bulk_obj@gsea$H, split = ".sign", label_format = 60) + ggplot2::facet_grid(~.sign)

```

### Gene Set Variation Analysis (GSVA)

```{r gsva-analysis, fig.width=6, fig.height=5}
# Run GSVA analysis -- add gsva results and diff results of gsva results
bulk_obj <- gsva(bulk_obj, geneSet_name = "H")

# # Run GSVA analysis with common used genesets
# bulk_obj <- gsva(bulk_obj, geneSet_name = c("H", "C2_CP:KEGG_LEGACY", "C5_GO:BP", "C5_GO:CC", "C5_GO:MF") )

# View bulk_obj--gsva results add to gsva slot
bulk_obj
# head(bulk_obj@gsva$H)

# Visualize GSVA results
# Bar plot
bar(bulk_obj, genes = "HALLMARK_INFLAMMATORY_RESPONSE", data = "H")

# Heatmap
hmap(bulk_obj, data = "H", genes = 20)

# Volcano plot
volcano(bulk_obj, data = "H")

# MA plot
MA(bulk_obj, data = "H")

```

## 📚 Documentation

For detailed tutorials and examples, please visit our [documentation website](https://hui950319.github.io/BulkS4/):

- [Getting Started Guide](https://hui950319.github.io/BulkS4/articles/getting-started.html)
- [Differential Expression Analysis](https://hui950319.github.io/BulkS4/articles/differential-analysis.html)
- [Visualization Tutorial](https://hui950319.github.io/BulkS4/articles/visualization.html)
- [Function Reference](https://hui950319.github.io/BulkS4/reference/index.html)

## 🤝 Contributing

We welcome contributions! Please see our [Contributing Guidelines](https://github.com/HUI950319/BulkS4/blob/master/.github/CONTRIBUTING.md) for details.

## 📄 License

This project is licensed under the MIT License - see the [LICENSE](LICENSE) file for details.

## 📞 Contact

- **Author**: Hui Ouyang
- **Email**: [your-email@example.com]
- **GitHub**: [@HUI950319](https://github.com/HUI950319)
- **Issues**: [GitHub Issues](https://github.com/HUI950319/BulkS4/issues)

## 📖 Citation

If you use BulkS4 in your research, please cite:

```
Ouyang, H. (2024). BulkS4: S4 Classes for Bulk RNA-seq Data Analysis. 
R package version 0.1.1. https://github.com/HUI950319/BulkS4
``` 